### 2021-03-22 manuscript/v5_submission_revision/replyreviewer.Rmd

The fitted vs residual plot shows general good model adequacy, since the residuals are almost evenly distributed around 0 for all fitted values. This means that the negative binomial model is adequate and there is no general non-linear trend in the predictors collectively (otherwise the black line would show a clear trend down/up/strongly curved).

The plot for individual predictors versus residuals does not reveal strong non-linear effects for individual predictors. The LOESS curves for COVID-19 burden and Mobility Principal Component 3 wiggle upwards for higher values, but these are very few. In the majority, curves are horizontal and don't imply model problems with non-linearity.

The distribution of the quantile residuals is a very good fit to a normal distribution, which indicates that the negative binomial model is a good model for the data. (For generalized linear models, QQ-plots of Pearson residuals are not meaningful, since a GLM actually makes no assumption about the distribution of errors.)

We did not model "time" explicitly as a variable. The residual plot with residuals ordered by time shows that we can assume independence of observations over time within the model, since the residuals are independent and identically distributed (iid) around 0 for every date. In reality, observations on subsequent days are not independent, but the data does not show a strong temporal trend in residuals.

### 2021-03-22 manuscript/v4_medrxiv_revision/results.Rmd
update all fixes. final manuscript for submission plos

```{r effectsrange, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Relative causal effects of exposures"}
knitr::include_graphics(rep('figures/f_exposure_vs_effect.eps'))
```

@@

Within our model, we observed effects of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate")), 2)` percent. On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate")), 2)` percent. We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to more cases). In perspective (Figure \@ref(fig:effectsrange)), with temperature we expect an increase by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate"))^(0-20)), 1)` percent at a daily average temperature of $0^{\circ}C$ compared to a day with $20^{\circ}C$. For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate"))^(10-0)), 1)` percent compared to a day with no precipitation.

@@

For all exposures, our analysis pipeline opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable "nursing homes", where the minimal adjustment set had a lower AIC. We found that there were no valid adjustment sets for the non-identifiable variables turnout and right-wing populist party votes.

>>>>

```{r effectsrange, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Relative causal effects of exposures"}
knitr::include_graphics(rep('figures/f_exposure_vs_effect_klein.eps'))
```

@@

Within our model, we observed effects of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate")), 2)` percent. On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate")), 2)` percent. We observe effects for humidity and wind as well (higher humidity and stronger wind leading to more cases). In perspective (Figure \@ref(fig:effectsrange)), with temperature we expect an increase by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate"))^(0-20)), 1)` percent at a daily average temperature of $0^{\circ}C$ compared to a day with $20^{\circ}C$. For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate"))^(10-0)), 1)` percent compared to a day with no precipitation.

@@

For all exposures, our analysis pipeline opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable "nursing homes", for which the minimal adjustment set had the lowest AIC. We found that there were no valid adjustment sets for the non-identifiable variables turnout and right-wing populist party votes.

### 2021-03-22 manuscript/v5_submission_revision/manuscript_tex4submission.tex
update all fixes. final manuscript for submission plos

We collected and aggregated data on reported COVID-19 cases, regional
socio-demographic factors, weather, and general mobility on district and
state level in Germany for the period of 15 February 2020 to 8 July
2020. Our observation period for the outcome consisted of all dates from
20 February 2020 to 8 July 2020 (\(T=140\)), since we used a lag of 5
days for all confounders. We did not exclude any states or districts
(\(K=401\)). We analyzed the daily reported number of new cases as
outcome (\(K\cdot T=56{\ }140\) observations). The set of possible
predictors was derived from our causal DAG (see Table~\ref{tab:model-vars} and Fig~\ref{fig:dag-covid-19}). Due to
modelling and data limitations, some of the predictors were unobserved
or were modelled as a construct consisting of several variables. For our causal graph analysis of effects, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).

@@

All continuous variables but the outcome ``Reported new cases of COVID-19'' and the offset ``Active cases'' were centered and scaled by one standard deviation for numerical stability, while we left binary variables as-is. After estimating the effects of variables, we re-scaled continuous variables' effects to their original scale. Additionally for mobility variables, we re-transformed the effects of the principal components to the original mobility variables. Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (see Section Results) which means that we assumed that their effects on the outcome will be visible after 5 days.

@@

We used a directed acyclic graph as a graphical representation of the
hypothesized causal reasoning that leads to exposure to the SARS-CoV-2
virus, onset of COVID-19, and finally reports of COVID-19 cases. We use the terms ``causal effect'' or ``causal relationship'' for effect estimates that are based on this causal graph framework. Every
node \(v_i\) in the graph is the graphical representation of an observed
or unobserved variable \(x_i\), a directed edge \(e_{ij}\) is an arrow
from node \(v_i\) to \(v_j\) that implies a direct causal relationship
from variable \(x_i\) onto variable \(x_j\). The set of all nodes is
denoted by \(V\), the set of all edges by \(E\), as such, the complete
DAG is the tuple \(G=(V,E)\). The seminal works of Spirtes and Pearl
\cite{spirtes2000causation, pearl2009causality} introduce the theory of
causal analysis, do-calculus, and how to analyze a DAG to estimate the
total or direct causal effect from a variable \(x_i\) onto a variable
\(x_j\). The direct effect is the effect associated with the edge
\(e_{ij}\) only (if it exists), while the total effect takes indirect
effects via other paths from \(v_i\) to \(v_j\) into account, too. Here
we estimated total effects only, since most of our variables were not
hypothesized to have a direct effect on the \emph{reported} number of
new COVID-19 cases. In contrast to prediction tasks, where one would
include all variables available, it is actually ill-advised to use all
available variables to estimate causal effects, due to introducing bias
by adjusting for unnecessary variables within the causal DAG. This is
why we need to identify a valid set of necessary variables (an
adjustment set) to estimate the proper causal effect
\cite{pearl2009causality}. The ``minimal adjustment set''
\cite{greenland1999causal} is a valid adjustment set of variables that
does not contain another valid adjustment set as a subset. However,
identifying a minimal adjustment set might not be enough to reliably
estimate the causal effect. Thus, we identified the ``optimal adjustment
set'' \cite{henckel2019graphical} as the set of variables which is a
valid adjustment set while having the lowest AIC.

@@

Within our model, we observed effects of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.95 percent. On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.21 percent. We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to more cases). In perspective (Fig~\ref{fig:effectsrange}), with temperature we expect an increase by approximately 21 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\). For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 12.8 percent compared to a day with no precipitation.

@@

For all exposures, our analysis pipeline opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable ``nursing homes'', where the minimal adjustment set had a lower AIC. We found that there were no valid adjustment sets for the non-identifiable variables turnout and right-wing populist party votes.

@@

Furthermore, our analyses support the effectiveness of non-pharmaceutical interventions. To a lesser extent, the adverse effects of some socio-demographic factors might help to identify areas that are at higher risk of local COVID-19 outbreaks.

>>>>

We collected and aggregated data on reported COVID-19 cases, regional
socio-demographic factors, weather, and general mobility on district and
state level in Germany for the period of 15 February 2020 to 8 July
2020. Our observation period for the outcome consisted of all dates from
20 February 2020 to 8 July 2020 (\(T=140\)), since we used a lag of 5
days for all confounders. We did not exclude any states or districts
(\(K=401\)). We analyzed the daily reported number of new cases as
outcome (\(K\cdot T=56{\ }140\) observations). The set of possible
predictors was derived from our causal DAG (see Table~\ref{tab:model-vars} and Fig~\ref{fig:dag-covid-19}). Due to
modelling and data limitations, some of the predictors were unobserved
or were modelled as a construct consisting of several variables. For our causal graph analysis, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).

@@

All continuous variables but the outcome ``Reported new cases of COVID-19'' and the offset ``Active cases'' were centered and scaled by one standard deviation for numerical stability, while we left binary variables as-is. After estimating the effects of variables, we re-scaled continuous variables' effects to their original scale. Additionally for mobility variables, we re-transformed the effects of the principal components to the original mobility variables. Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (optimal lag found by cross-validation) which means that we assumed that their effects on the outcome will be visible after 5 days.

@@

We used a directed acyclic graph as a graphical representation of the
hypothesized causal reasoning that leads to exposure to the SARS-CoV-2
virus, onset of COVID-19, and finally reports of COVID-19 cases. We use the terms ``causal effect'' or ``causal relationship'' for effect estimates that are based on this causal graph framework. Every
node \(v_i\) in the graph is the graphical representation of an observed
or unobserved variable \(x_i\), a directed edge \(e_{ij}\) is an arrow
from node \(v_i\) to \(v_j\) that implies a direct causal relationship
from variable \(x_i\) onto variable \(x_j\). The set of all nodes is
denoted by \(V\), the set of all edges by \(E\), as such, the complete
DAG is the tuple \(G=(V,E)\). The seminal works of Spirtes and Pearl
\cite{spirtes2000causation, pearl2009causality} introduce the theory of
causal analysis, do-calculus, and how to analyze a DAG to estimate the
total or direct causal effect from a variable \(x_i\) onto a variable
\(x_j\). The direct effect is the effect associated with the edge
\(e_{ij}\) only (if it exists), while the total effect takes indirect
effects via other paths from \(v_i\) to \(v_j\) into account, too. Here
we estimated total effects only, since most of our variables were not
hypothesized to have a direct effect on the \emph{reported} number of
new COVID-19 cases. In contrast to prediction tasks, where one would
include all variables available, it is actually ill-advised to use all
available variables to estimate causal effects, due to introducing bias
by adjusting for unnecessary variables within the causal DAG. This is
why we need to identify a valid set of necessary variables (an
adjustment set) to estimate the proper causal effect
\cite{pearl2009causality}. The ``minimal adjustment set''
\cite{greenland1999causal} is a valid adjustment set of variables that
does not contain another valid adjustment set as a subset. However,
identifying a minimal adjustment set might not be enough to reliably
estimate the causal effect. Thus, we identified the ``optimal adjustment
set'' \cite{henckel2019graphical} as the set of variables which is a
valid adjustment set while having the lowest Akaike information criterion (AIC).

@@

Within our model, we observed effects of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.95 percent. On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.21 percent. We observe effects for humidity and wind as well (higher humidity and stronger wind leading to more cases). In perspective (Fig~\ref{fig:effectsrange}), with temperature we expect an increase by approximately 21 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\). For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 12.8 percent compared to a day with no precipitation.

@@

For all exposures, our analysis pipeline opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable ``nursing homes'', for which the minimal adjustment set had the lowest AIC. We found that there were no valid adjustment sets for the non-identifiable variables turnout and right-wing populist party votes.

@@

Furthermore, our analyses strongly support the effectiveness of non-pharmaceutical interventions. To a lesser extent, the adverse effects of some socio-demographic factors might help to identify areas that are at higher risk of local COVID-19 outbreaks.
